Best Practices in Latent Class Analysis using Free Open Source Software

### 2022-11-22 manuscript.Rmd
Update November 23
- updated the literature list
- added use case examples
- corrected spelling/grammar mistakes
- adjusted spacing, also in the tutorial

*** things to resolve:
1) Caspar should add Well-being compass study to Zotero and update the in-text citation for that use case
2) the plot_density function doesn't work, also the example in the documentation;
I would like to show how to use it in class enumeration section of the tutorial
3) final proof-reading, anything to add?

LCA is like confirmatory factor analysis, except that the continuous
latent variable is substituted with a categorical latent variable. Both
assume that the indicators are related, and these relations can be
represented by the underlying latent variable. One difference between
the two techniques is that factor analysis can be considered as a way to
group observed *variables* into latent constructs, and LCA groups
*individuals* into classes.While factor analysis seeks to describe the
relations between variables, LCA seeks to describe each unobserved group
and the differences between them. In line with this distinction, LCA
(mixture models) is referred to as a "person-centered" technique, and
factor analysis as a "variable-centered" technique
[@nylund-gibson_ten_2018; @masyn_latent_2013].

@@

In factor analysis, the latent variable is continuous, so the relation
between indicators is defined by the communalities between them, and the
latent variable represents the error corrected approximation of the
sample. While in LCA, the relation between indicators is define by how
well they separate the classes, as a multinomial regression between
indicators and latent variable [@masyn_latent_2013]

When the focus is on each individual's estimated class membership, LCA
can be thought of as a type of clustering algorithm. In line with this
perspective, LCA is sometimes described as "model-based clustering"
[@hennig_handbook_2015; @scrucca_mclust_2016]. Many clustering
algorithms apply some recursive splitting algorithm to the data. By
contrast "model-based" clustering refers to the fact that LCA estimates
cluster membership based on a parametric model. Specifically, the
posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data
under given the class-specific model. For these differences, LCA is a clustering method, that corrects for (random) measurement error, while clustering algorithms do not correct for. This is something LCA has in common with factor analysis, as both model an underlying latent variable, which accounts for the communalities between indicators and correcting for measurement error. 0

@@

The technique known as *latent profile analysis (LPA)* is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for, the covariances/correlations between indicators are
assumed to be zero. This can be conceived of as a restricted mixture
model with covariances fixed to zero. Stating that the covariances/correlations are accounted by the latent variable. In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes [@collins_latent_2009]. As an example, a LPA of ocean
plastic particles found two classes of particles based on length and
width: a class of smaller particles with a high correlation between
length and width, meaning that these particles were approximately round
or square in shape, and a class of larger particles with a low
correlation between length and width, meaning that these particles were
heterogenous in shape. From a theoretical perspective, this makes sense,
because the smaller particles have been ground down to a more uniform
shape by the elements.
<!-- DA: what is the source for plastic particles? -->

It is also possible to estimate a mixture model based on latent
indicators. This means that, within each class, one or more continuous
latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated based on these
continuous latent variables. A common application of this approach is in
longitudinal research, where the indicators reflect one construct
assessed at different time points. Examples of this approach include
*growth mixture models* (GMM) and *latent class growth analyses* (LCGA).
These techniques estimate a latent growth model to describe grouped
trajectories over time. The growth mixture model is a latent class model
where the parameters that indicate class membership are the intercepts
and variances, and typically covariances of the latent growth variables,
e.g., a latent intercept and slope. This technique assumes that
individuals within a class can have heterogenous trajectories. If the
variance of the growth parameters is fixed to zero, it is known as a
latent class growth analysis. This latter approach assumes that all
individuals within a class share the same identical trajectory, and that
any variance in the indicators not explained by the class-specific
latent trajectories is due to residual error variance.
<!-- MGV: some reference for this description -->

@@

<!-- Caspar will check if I can replace them with examples from Caspar's published papers. -->

There are several use cases for which LCA methods are suitable. One
example is to test a theory that postulates the existence of a
categorical latent variable. For example, *identity status theory*
posits that, at any given point in time, adolescents reside in one of
four identity statuses. Latent class analysis can be used to identify
these four statuses based on observed indicators (e.g., self-reported
identity exploration and commitment). If results indicate that the data
are better described by a different number of classes, or that the
four-class solution does not correspond to the predicted pattern of
responses on the indicators, then the theory may be called into
question.

Another use case is unsupervised learning: when the goal is to restore
unobserved class membership based on observed indicators, or to classify
individuals. For example, a mixture model can be used as a diagnostic
aid when several clinical indicators can be used to distinguish between
a fixed number of physical [@baughman_mixture_2006], mental
[@wu_abuse_2011] health problems, or health outcomes after recovering from cancer [@hardy_exploring_2018]. The example of shoe size is a
rudimentary illustration of this type of application.

LCA can be used as a descriptive analysis where a researcher wishes to
describe their sample and identify a few prototypes based on many
variables. As an example, if a survey among all Dutch academics was
carried out with the goal of bringing about a funding reform, LCA could
be used to discover the types of publications that get funding.

With LCA, our goal could be to inductively identify the number of
classes. For instance, if we believe that a variable represents a group,
but don't know how many groups there are, LCA may be an appropriate
technique to answer this question. For example, @hopfer_social_2014
studied substance use, sexual behavior, and mental health status of
urban population in Winnipeg, Canada. The underlying assumption was that
there were different risk profiles, but their number was not known. From
a collection of indicators, LCA provided evidence that there are four
distinct risk profiles in the Winnipeg area.

Another application of LCA is to classify individuals. In a peer
harassment study, @giang_using_2008 used LCA to classify over 2,000
sixth grade students into aggressor and victim latent classes. The
five-class solution comprised of classes of victims, aggressors, and
socially adjusted students. For instance, it revealed that there were
two types of victims: highly-victimized aggressive-victims and
highly-aggressive aggressive-victims.

LCA is also appropriate when we wish to identify indicators that capture
classes well. High quality indicators will be strongly related to the
latent variable and will lead to good class separation. This
relationship of high quality indicators to the latent variable is
reflected in very high or very low conditional response probabilities.
For a simulation study exploring the effects of indicator quality on
LCA, see @geiser_is_2014. Therefore, one could use conditional response
probabilities for each item to assess its quality with regards to how
well it helps separate the latent classes. From this, a theory about the
selection of indicators could be informed.

An extension of LCA is that containing covariates which can be used to
predict class membership. In this approach, we not only model the latent
class variable based on indicators, but we also relate the class
membership to other explanatory variables (@vermunt_latent_2017). An
example of using covariates comes from @nozadi_moderating_2016 who
applied LCA to identify the probability of children's membership to an
anxiety class. The authors tested several covariates including
children's age, sex, and accuracy scores. Age and sex were not found to
be related to the children's latent class membership, hence these
covariates were excluded from the analysis. Accuracy scores, however,
were related to probabilities of being in anxiety and attention-anxiety
classes. Therefore this covariate was kept as a valuable predictor.

When our interest is the prediction of one or more outcomes, LCA can be
used to construct latent classes as categorical predictors.
@lanza_latent_2013 demonstrated how LCA can be used to classify
adolescents into depression classes, and subsequently these classes can
be used to predict smoking, grades, and delinquency. The study showed
that the outcomes predicted by class membership can be binary (regular
smoking), continuous (grades) or count (delinquency).

These mixture models can also be useful to evaluate quality of measurement across instrument. For example, employment status was collected across register and survey data, and there is disagreement between them. You could think that register data is without error, but it has been shown that is victim of measurement error as well. In this case mixture models was used to evaluate the underlying factor of "error corrected employment status", by using two indicators of the same question and identify random and systamtic measurement error in either indicator [@pavlopoulos_measuring_2015].

In addition to these applications, LCA can be used for dimensionality
reduction as the resulting groups summarize response patterns on a large
number of indicators. For example, @macgregor_symptom_2021 investigated
symptom profiles among injured U.S. military personnel. They used
fifteen dichotomous items from the Post-Deployment Health Assessment
survey as LCA indicators. Combinatorics informs us that fifteen
dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA as a
method of dimensionality reduction. A five class solution was found to
have the best fit according to both statistical criteria and clinical
interpretability.

<!-- CJ: (on the next paragraph)  -->

<!-- the multilevel LCA addresses violations of the assumptions of LCA. -->

<!-- That is not what this paragraph is about.  -->

<!-- This paragraph is about motivations for using LCA. -->

<!-- One motivation is if your data violates multivariate normality -->

<!-- because violating multivariate normality means that you have multiple groups -->

<!-- and you can use LCA to extract those. That is the only assumption  -->

<!-- we have to discuss here, we just need to explain it. -->

Finally, LCA can be used to deal with data which violate certain
assumptions. As discussed in the shoe size example, LCA can deal with
violations of normality. In fact, LCA assumes the population
distribution is a non-normal mixture of $K$ normal distributions, and it
can discover the value of $K$, i.e. generate a $K$-class solution.

LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous of categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question.

# Best Practices

@@

While we concur that investigating missingness is due diligence, it is
important to emphasize that missingness is adequately handled by default
in many software packages for LCA, such as Mplus, Latet GOLD, and `OpenMx` which is
the backend of `tidySEM`. These packages use Full Information Maximum
Likelihood (FIML) estimation, which makes use of all available
information without imputing missing values. FIML is a best-practice
solution for handling missing data; on par with multiple imputation when the estimation model includes the same variables as the imputation model
[@lee_comparison_2021]. FIML estimation assumes that missingness is
either MCAR or MAR. Thus, one would typically proceed with FIML,
regardless of the outcome of an MCAR test. Although FIML does not, by
default, handle missingness in exogenous variables -- all indicator
variables in LCA are endogenous, so this is not a concern.
<!-- are you sure Mplus accounts for missing data automatically when the indicators are categorical? I know that is not the case with CFA/SEM models -->

@@

Fit indices typically used for determining the optimal number of classes
include the Akaike Information Criterion (AIC) and the Bayesian
Information Criterion (BIC). Both information criteria are based on the
-2\*log-likelihood (which is lower for better fitting models), and add a
penalty for the number of parameters (thus incentivizing simpler
models). This helps balance model fit and model complexity. The lower
the value of an information criterion, the better the overall fit of the
model.

The general objective of the Information Criteria (IC) is to evaluate the model's out-of-sample predictive accuracy, so adjusting for over fitting. Fit measures like $R^2$ evaluate the in-sample predictive accuracy, meaning that they evaluate the models ability to predict the observed outcomes based on the same data that was use to build up the model. These metrics are positively bias, meaning that will present better model fit than it has in reality. While IC corrects for this positive bias by evaluating the models accuracy approximating the out-of-sample predictive accuracy, meaning that it is the ability to predict the outcome for observations that are not part of the training model. Ideally we would estimate the model a lot of time with less observations and predict their scores, but IC's approximate this by their different penalty metrics. For example $R^2$ will increase even if an added predictors are unnecessary, while IC's will show worst fit when a predictor (complexity) is unnecessary [@mcelreath_statistical_2020]

@@

A disadvantage of the IC's is that we do not have standard errors for them, so we only have the absolute values without a measure of their variability. So, the difference between models IC can be very small. Still indicating that the model with the lower value is "better", but if this difference is  very small can considered them "functionally equal", and you should take into consideration the interpretability of the model.

@@

`$individual` is the individual posterior probability matrix, with
dimensions n (number of cases in the data) x k (number of classes). Additionally it includes the `predicted` class in function of the highest predicted probability.
Individual class probabilities and/or predicted class are often useful for researchers who wish to do follow up analyses.

@@

A problem which can occur is that of inadmissible solutions. With binary
indicators, LCA is modelling a cross-table with all the predictors. The
problem with such cross-tables is that they will often contain empty
cells, i.e. combinations of responses that never occur together. This
problem is reflected by extreme conditional item probabilities (as in
exactly 0 or 1). Such boundary parameter estimates could indicate that
the solution is invalid [@geiser_data_2012]. Boundary parameter
estimates can also happen with continuous indicators. For instance, if
we have a zero-inflated normal distribution and a two class solution,
one class might have the mean of zero and its standard deviation cannot
be determined since there is little variance. This too could be a sign
of an invalid solution, warn that too many classes were extracted, or
indicate a local optimum [@geiser_data_2012].

@@

First, in class enumeration, we want to compare several competing class
solutions. This can be done by means of the AIC and BIC. Here, we
suggest using an elbow plot with the $K-class\ solution$ on the X-axis,
and information criteria value on the Y-axis. We show an implementation
of such an elbow plot in our Tutorial subsection.
<!-- For continuous variables, density plots can aid class enumeration. -->
<!-- DA: Do we have density plots implemented in tidySEM?, yes, it is plot_density  -->
<!-- If so, we should add this example to the tutorial. -->

@@

In this example, we use *data_mix_ordinal*, a simulated data for mixture
modelling with ordinal indicators. This dataset is built-in to
`tidySEM`. For more information about the dataset, type
`?tidySEM::data_mix_ordinal` into the R console after loading the
`tidySEM` using library().

@@

```{r fitting_LCA, include = TRUE, eval=F}
set.seed(123) # setting seed 
res <- mx_lca(data = df, classes = 1:4) # fitting LCA - 1 to 4 classes
```

Notice that the model was run 10 times for each $K$ class solution. This because the functions runs a total of 10 **extra** times the model. This way we can choose the best fitting model across all the solutions. In this examples we see that all models converged without issues.

@@

As `tidySEM::mx_lca()` was run for several models at the same time, the resulting object `res` is a `mixture_list` class where each $K$ class solution is an object in the list, and each of them are `OpenMx` models (`MxModel`).

@@

```{r, echo=F}
Selection <- c("Name", "LL", "Parameters", "AIC", "BIC", "prob_min", "prob_max",
          "n_min", "n_max")
Description <- c("the $K$-class solution",
                 "the -2*log-likelihood of each model",
                 "Number of estimated parameters in each model",
                 "the Akaike Information Criterion value",
                 "the Bayesian Information Criterion value",
                 "the lowest posterior class probability by most likely class membership",
                 "the highest posterior class probability by most likely class membership",
                 "the lowest class proportion based on the posterior class probabilities",
                 "the highest class proportion based on the posterior class probabilities")
selection <- cbind(Selection, Description)
knitr::kable(selection, format = 'pipe', caption = "Selection of Fit Indices and Classification Diagnostics")
```

@@

```{r plot_prob, include = TRUE, eval=F}
plot_prob(res[[3]]) # visualizing the response patterns for the final class solution
table_prob(res[[3]]) # tabulating the response patterns for the final class solution
```

@@

```{r extract_Post_Class_Prob, include = TRUE, eval=F}
probs <- class_prob(res[[3]]) # extracting the posterior class probabilities of the final class solution
probs$mostlikely.class # posterior class probabilities by most likely class membership
probs$individual # individual posterior class probabilities
```

>>>>

LCA is like confirmatory factor analysis, except that the continuous
latent variable is substituted with a categorical latent variable. Both
assume that the indicators are related, and these relations can be
represented by the underlying latent variable. One difference between
the two techniques is that factor analysis can be considered as a way to
group observed *variables* into latent constructs, and LCA groups
*individuals* into classes. While factor analysis seeks to describe the
relations between variables, LCA seeks to describe each unobserved group
and the differences between them. In line with this distinction, LCA
(mixture models) is referred to as a "person-centered" technique, and
factor analysis as a "variable-centered" technique
[@nylund-gibson_ten_2018; @masyn_latent_2013].

@@

In factor analysis, the latent variable is continuous, so the relationships
between indicators are defined by the communalities between them. Here, the
latent variable represents the error corrected approximation of the
sample. In LCA, by contrast, the relation between indicators is defined by how
well they separate the classes, as a multinomial regression between the
indicators and the latent variable [@masyn_latent_2013].

When the focus is on each individual's estimated class membership, LCA
can be thought of as a type of clustering algorithm. In line with this
perspective, LCA is sometimes described as "model-based clustering"
[@hennig_handbook_2015; @scrucca_mclust_2016]. Many clustering
algorithms apply some recursive splitting algorithm to the data. By
contrast "model-based" clustering refers to the fact that LCA estimates
cluster membership based on a parametric model. Specifically, the
posterior class probability that an individual belongs to a latent class
can be computed from the likelihood of that individual's observed data
under given the class-specific model. For these differences, 
LCA is a clustering method, that corrects for (random) measurement error, 
while clustering algorithms do not correct for. This is something LCA has 
in common with factor analysis, as both model an underlying latent variable, 
which accounts for the communalities between indicators 
and correcting for measurement error.

@@

The technique known as *latent profile analysis (LPA)* is a special case
of the mixture model, which assumes conditional independence of the
indicators. Conditional independence means that, after class membership
is accounted for, the covariances/correlations between indicators are
assumed to be zero. This can be conceived of as a restricted mixture
model with covariances fixed to zero. Stating that the covariances/correlations
are accounted by the latent variable. In some cases, such constraints
will be inappropriate; for instance when the cohesion between indicators
is expected to differ between classes [@collins_latent_2009]. As an example,
a LPA of ocean plastic particles found two classes of particles based on length
and width: a class of smaller particles with a high correlation between
length and width, meaning that these particles were approximately round
or square in shape, and a class of larger particles with a low
correlation between length and width, meaning that these particles were
heterogenous in shape [@alkema_maximizing_2022]. From a theoretical perspective, this makes sense,
because the smaller particles have been ground down to a more uniform
shape by the elements.

It is also possible to estimate a mixture model based on latent
indicators. This means that, within each class, one or more continuous
latent variables are estimated based on the observed indicators.
Categorical latent variable membership is then estimated based on these
continuous latent variables. A common application of this approach is in
longitudinal research, where the indicators reflect one construct
assessed at different time points. Examples of this approach include
*growth mixture models* (GMM) and *latent class growth analyses* (LCGA) 
[@jung_introduction_2008]. These techniques estimate a latent growth model 
to describe grouped trajectories over time. The growth mixture model is a 
latent class model where the parameters that indicate class membership are the 
intercepts and variances, and typically covariances of the latent growth 
variables, e.g., a latent intercept and slope. This technique assumes that
individuals within a class can have heterogenous trajectories. If the
variance of the growth parameters is fixed to zero, it is known as a
latent class growth analysis. This latter approach assumes that all
individuals within a class share the same identical trajectory, and that
any variance in the indicators not explained by the class-specific
latent trajectories is due to residual error variance.

@@

There are several use cases for which LCA methods are suitable. One
such application is to test a theory that postulates the existence of a
categorical latent variable. For instance, *identity status theory*
posits that, at any given point in time, adolescents reside in one of
four identity statuses [@marcia_development_1966; @marcia_identity_1980]. 
LCA can be used to test whether these four statuses
can be recovered based on observed indicators (e.g., self-reported
identity exploration and commitment). If the results indicate that the data
are better described by a different number of classes, or that the
four-class solution does not correspond to the predicted pattern of
responses on the indicators, then the theory may be called into
question. A recent study by @maene_perceived_2022 applied this principle to 
test whether identities of Belgian adolescents with migration background can be 
summarized as those relating to their heritage, national and regional identity.
A LCA revealed three identity profiles: full integration (high belonging to 
heritage, national and regional identity), national integration (high heritage
and national identity with low regional identity), and weak separation
(high heritage identity with neutral attitudes towards the nation and region).

Another use case is unsupervised learning: when the goal is to restore
unobserved class membership based on observed indicators, or to classify
individuals. For example, a mixture model can be utilized as a diagnostic
aid when using several clinical indicators to distinguish between
a fixed number of physical [@baughman_mixture_2006], mental health
[@wu_abuse_2011] problems, or health outcomes after recovering from
cancer [@hardy_exploring_2018]. The example of shoe size is a rudimentary
illustration of this type of application. As another example, a recent study 
identified two suicide risk profiles in military service members and veterans
[@gromatsky_characteristics_2022].  One class was defined by higher-risk
history and another by lower-risk history. The higher-risk group comprised of
individuals with greater intent to die using methods with higher lethality,
and the lower-risk group was characterized by greater ambivalence towards death
and usage of lower lethality methods.

A related use case is where class-membership is known but imperfectly measured.
Mixture models can also be useful to evaluate the quality of measurement
across instruments. For example, a recent study had employment status 
information collected across register and survey data. There was disagreement 
between the two data sources. While it might be tempting to assume that the
register data is free of error, this has been shown to be false.
The authors used a mixture model to evaluate the underlying factor of 
"error corrected employment status" by using two indicators of the same question
and identifying random and systematic measurement error in both indicators 
[@pavlopoulos_measuring_2015].

LCA can be used as a descriptive analysis where a researcher wishes to
describe their sample and identify a few prototypes based on many
variables. As an example, if a survey among all Dutch academics was
carried out with the goal of bringing about a funding reform, LCA could
be used to discover the types of publications that get funding.
Recently, @alkema_maximizing_2022 studied plastic debris in the Atlantic Ocean.
Prior to this study, ways to categorize ocean microplastics in terms of length,
width and polymer type were rather arbitrary, and assumed that these properties
were independent. Also, there was a lack of information regarding variability 
within these categories. The authors used finite Gaussian mixture modeling to 
build a classification system for ocean microplastic particles based on a set 
of measured characteristics. With regards to particle dimensions, two classes
emerged: one with smaller fragments whose width and height were highly 
correlated, and another consisting of larger fragments with low correlation 
between particle width and height.

With LCA, our goal could be to inductively identify the number of classes. 
For instance, if we believe that a variable represents a group,
but do not know how many groups there are, LCA may be an appropriate
technique to answer this question. For example, @hopfer_social_2014
studied substance use, sexual behavior, and mental health status of
urban population in Winnipeg, Canada. The underlying assumption was that
there were different risk profiles, but their number was not known. 
From a collection of indicators, LCA provided evidence that there are four
distinct risk profiles in the Winnipeg area. Another study investigated 
adolescent empathy development and its influence on parent-child conflict 
[@van_lissa_divergence_2015]. The data was collected over six years. 
Using LCGA, adolescents with similar empathy development trajectories 
were grouped into: high-, average-, and low-empathy classes.

Another application of LCA is to classify individuals. In a peer harassment 
study, @giang_using_2008 used LCA to classify over 2,000 sixth grade students 
into aggressor and victim classes. The five-class solution comprised of 
classes of victims, aggressors, and socially adjusted students. 
The study revealed that there were two types of victims: highly-victimized 
aggressive-victims and highly-aggressive aggressive-victims.
Another example is the "Well-being compass" study by van Lissa, currently 
a preprint. The study was conducted in collaboration with a mental health care
facility and focused on care providers. Latent class analysis has identified 
four different risk profiles for care providers. Not only was class membership 
estimated for the participants of the original study, but the class solution 
was used as a risk assessment aid for new care providers, 
so they could be more adequately supported by mental health care professionals.
<!-- DA: this study is not in zotero yet, I lack information (title, authors, etc.) to enter it, @Caspar could you enter this to Zotero and adjust the in-text citation-->

LCA is also appropriate when we wish to identify indicators that capture
classes well. High quality indicators will be strongly related to the
latent variable and will lead to good class separation. This
relationship of high quality indicators to the latent variable is
reflected in very high or very low conditional response probabilities.
For a simulation study exploring the effects of indicator quality on
LCA, see @geiser_is_2014. Therefore, one could use conditional response
probabilities for each item to assess its quality with regards to how
well it helps separate the latent classes. From this, a theory about the
selection of indicators could be informed.
An example of such use of LCA is a study by @tsaousis_measurement_2020 who
studied differential item functioning using LCA. The study focused 
on university admission tests of Saudi Arabian students. LCA has revealed three
latent classes: high-, average-, and low-scorers. Subsequently, the authors
used the Multiple Indicator Multiple Causes model to identify both uniform
and non-uniform differential item functioning. The main finding was that gender
was a potential source of differential item functioning for latent class
indicators. Therefore, gender should be included as a covariate when fitting 
a LCA model in order to obtain unbiased estimates.

An extension of LCA is that containing covariates which can be used to
predict class membership. In this approach, we not only model the latent
class variable based on indicators, but we also relate the class
membership to other explanatory variables (@vermunt_latent_2017). An
example of using covariates comes from @nozadi_moderating_2016 who
applied LCA to identify the probability of children's membership to an
anxiety class. The authors tested several covariates including
children's age, sex, and accuracy scores. Age and sex were not found to
be related to the children's latent class membership, hence these
covariates were excluded from the analysis. Accuracy scores, however,
were related to probabilities of being in anxiety and attention-anxiety
classes. Therefore this covariate was kept as a valuable predictor.
Another example comes from @van_lissa_divergence_2015 who included sex 
as a predictor of the developmental trajectories and class membership in 
the previously discussed adolescent empathy development study. The rationale
was that adding sex as a covariate would account for known sex differences 
in empathy development, and lead to a model with lower bias.

When our interest is the prediction of one or more outcomes, LCA can be used 
to construct latent classes as categorical predictors. @lanza_latent_2013 
demonstrated how LCA can be used to classify adolescents into depression 
classes, and subsequently these classes can be used to predict smoking, grades,
and delinquency. The study showed that the outcomes predicted by class 
membership can be binary (regular smoking), continuous (grades) or 
count (delinquency). In the previously discussed study of Belgian adolescent
identities, @maene_perceived_2022 first constructed the latent variable
representing different multiple identification strategies. Subsequently, 
the most likely class membership based on posterior class probability was
used as an "observed" categorical variable for prediction of depressive symptoms.

In addition to these applications, LCA can be used for dimensionality
reduction as the resulting groups summarize response patterns on a large
number of indicators. For example, @macgregor_symptom_2021 investigated
symptom profiles among injured U.S. military personnel. They used
fifteen dichotomous items from the Post-Deployment Health Assessment
survey as LCA indicators. Combinatorics informs us that fifteen
dichotomous items have $2^{15}$ or $32,768$ unique symptom combinations.
Perhaps for this reason, @macgregor_symptom_2021 incorporated LCA as a
method of dimensionality reduction. A five class solution was found to
have the best fit according to both statistical criteria and clinical
interpretability. In the longitudinal adolescent empathy development study, 
@van_lissa_divergence_2015 first investigated heterogeneity in empathy 
development. This was done through identification of classes of individuals 
characterized by similar developmental trajectories. The authors used LCGA
as a dimensionality reduction technique where 467 individual trajectories
were summarized into three empathy classes. Such reduction can help researchers 
gain a better perspective of the grand patterns in the data.

Finally, LCA can be used to deal with data which violate certain
assumptions. As discussed in the shoe size example, LCA can deal with
violations of normality. In fact, LCA assumes the population
distribution is a non-normal mixture of $K$ normal distributions, and it
can recover the value of $K$, i.e. generate a $K$-class solution.
For instance, @alkema_maximizing_2022 studied ocean plastic debris. 
Plastic particle width and height are non-normally distributed. The authors 
demonstrate that the non-normal distributions of particle dimensions are
due to the measurements being a mixture composed of several particle types, 
each with its own normal distribution.

<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question. -->
<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->
# Best Practices

@@

While we concur that investigating missingness is due diligence, it is
important to emphasize that missingness is adequately handled by default
in many software packages for LCA, such as Mplus, Latet GOLD, and `OpenMx` 
which is the backend of `tidySEM`. These packages use Full Information Maximum
Likelihood (FIML) estimation, which makes use of all available
information without imputing missing values. FIML is a best-practice
solution for handling missing data; on par with multiple imputation when the 
estimation model includes the same variables as the imputation model
[@lee_comparison_2021]. FIML estimation assumes that missingness is
either MCAR or MAR. Thus, one would typically proceed with FIML,
regardless of the outcome of an MCAR test. Although FIML does not, by
default, handle missingness in exogenous variables -- all indicator
variables in LCA are endogenous, so this is not a concern.
<!-- are you sure Mplus accounts for missing data automatically when the indicators are categorical? I know that is not the case with CFA/SEM models -->
<!-- CJ: yes, it does -->

@@

Fit indices typically used for determining the optimal number of classes
include the Akaike Information Criterion (AIC) and the Bayesian
Information Criterion (BIC). Both information criteria are based on the
-2\*log-likelihood (which is lower for better fitting models), and a
penalty term for the number of parameters (thus incentivizing simpler
models). This helps balance model fit and model complexity. The lower
the value of an information criterion, the better the overall fit of the
model. 
XXXXXX
The general objective of information criteria (ICs) is to evaluate the model's 
out-of-sample predictive accuracy, thus adjusting it for overfitting. 
Fit measures such as $R^2$ evaluate the in-sample predictive accuracy, 
i.e. the model's ability to predict the observed outcomes based on the same 
data that was used to fit the model. In-sample metrics are positively biased, 
meaning that they overestimate the fit the model has in reality, i.e. when 
presented with new data. While common ways to overcome this bias is through
resampling methods such as cross-validation, or use of out-of-sample data such
as a test set, ICs correct for this positive bias by evaluating the model's 
accuracy in a way which approximates the out-of-sample predictive accuracy. 
ICs achieve this by applying different penalty metrics. For example, $R^2$ will 
increase even if an added predictor is fully redundant. ICs show preference for
simpler models and indicate worse fit when a predictor, or its added complexity 
is unnecessary [@mcelreath_statistical_2020].

@@

A disadvantage of ICs is that they are not accompanied by standard errors or 
any measure of variability, only their absolute value. The difference between 
two models' ICs can be rather small. The lower information criterion would 
still indicate the better model, but if the ICs difference is very small, 
two models can be considered "functionally equal". In such situations, the 
interpretability of the model should take priority.

@@

`$individual` is the individual posterior probability matrix, with
dimensions n (number of cases in the data) x k (number of classes). 
Additionally, it includes the `predicted` class as a function of the highest 
predicted probability. Individual class probabilities and/or predicted class 
are often useful to researchers who wish to carry out follow up analyses.

@@

A problem which can occur is that of inadmissible solutions. With binary
indicators, LCA is modeling a cross-table with all the predictors. The
problem with such cross-tables is that they will often contain empty
cells, i.e. combinations of responses that never occur together. This
problem is reflected by extreme conditional item probabilities (as in
exactly 0 or 1). Such boundary parameter estimates could indicate that
the solution is invalid [@geiser_data_2012]. Boundary parameter
estimates can also happen with continuous indicators. For instance, if
we have a zero-inflated normal distribution and a two class solution,
one class might have the mean of zero and its standard deviation cannot
be determined since there is little variance. This too could be a sign
of an invalid solution, warn that too many classes were extracted, or
indicate a local optimum [@geiser_data_2012].

@@

First, in class enumeration, we want to compare several competing class
solutions. This can be done by means of the AIC and BIC. Here, we
suggest using an elbow plot with the $K-class\ solution$ on the X-axis,
and information criteria value on the Y-axis. For continuous variables, 
density plots can aid class enumeration. We demonstrate how to implement 
both elbow and density plots in the Tutorial subsection of this paper.

@@

In this example, we use *data_mix_ordinal*, a simulated data for mixture
modeling with ordinal indicators. This dataset is built-in to
`tidySEM`. For more information about the dataset, type
`?tidySEM::data_mix_ordinal` into the R console after loading the
`tidySEM` using library().

@@

```{r fitting_LCA, include = TRUE, eval=F}
set.seed(123) # setting seed 
res <- mx_lca(data = df, classes = 1:4) # fitting LCA 1 to 4 class solutions
```

Notice that the model run 10 times for each $K$ class solution. 
This way we can choose the best fitting model across all the solutions. 
In this example, we see that all models converged without issues.

We can inspect the class of resulting objects.

@@

As `tidySEM::mx_lca()` run several models at a time, the resulting object `res` 
is a `mixture_list` class where each $K$ class solution is a list element, or
more specifically an `OpenMx` model (`MxModel`).

@@

```{r, echo=F}
Selection <- c(
    "Name", "LL", "Parameters", 
    "AIC", "BIC", "prob_min", 
    "prob_max", "n_min", "n_max")

Description <- c(
    "the $K$-class solution",
    "the -2*log-likelihood of each model",
    "Number of estimated parameters in each model",
    "the Akaike Information Criterion value",
    "the Bayesian Information Criterion value",
    "the lowest posterior class probability by most likely class membership",
    "the highest posterior class probability by most likely class membership",
    "the lowest class proportion based on the posterior class probabilities",
    "the highest class proportion based on the posterior class probabilities")

selection <- cbind(Selection, Description)

knitr::kable(selection, format = 'pipe', 
             caption = "Selection of Fit Indices and 
                        Classification Diagnostics")
```

@@

```{r plot_prob, include = TRUE, eval=F}
plot_prob(res[[3]]) # visualizing the response patterns for the final model
table_prob(res[[3]]) # tabulating the response patterns for the final model
```

@@

```{r extract_Post_Class_Prob, include = TRUE, eval=F}
probs <- class_prob(res[[3]]) # extracting the posterior class probabilities
probs$mostlikely.class # posterior probabilities by most likely class membership
probs$individual # individual posterior class probabilities
```
